---
title: ''
author: "Basil Seif"
date: "8/13/2019"
output:
  pdf_document: default
  html_document: default
---
In tackling this exercise, I decided to build a ridge regression model that assigns betas to each team based on home performance and road performance, with home team score margin for each game as the target variable. The final score for each team is the sum of those home and road performance coefficient values. The prediction matrix also includes the $NeutralSite$ binary variable, meaning that team performance for games played at neutral sites are baked into these home and away coefficient values. 

I decided to use ridge regression in order to minimize the feature set while also reducing collinearity between teams that played against each other more than average. 

My first step was to read in the data and create a reference table for team IDs and game IDs.

```{r}
####Read in Libraries####
library(data.table)
library(dplyr)
library(Hmisc)
library(glmnet)
library(tidyr)
library(ranger)


####Read in Data####
exampleRanks <- fread("ExampleRankings.csv", sep = ",", stringsAsFactors = F)
ACCGames <- fread("ACCGames1819.csv", sep = ",", stringsAsFactors = F)


####Cleaning Data####
ACCGames$GameDate <- as.Date(ACCGames$GameDate, format = "%m/%d/%Y")
teamIDRefs <- data.frame(team_name = sort(unique(c(ACCGames$AwayTeam, ACCGames$HomeTeam))),
                         team_id = c(1:length(sort(unique(c(ACCGames$AwayTeam, ACCGames$HomeTeam))))))
ACCGames <- merge(ACCGames, teamIDRefs, by.x = "AwayTeam", by.y = "team_name", all.x = T)
names(ACCGames)[which(names(ACCGames) == "team_id")] = "AwayTeamID"
ACCGames <- merge(ACCGames, teamIDRefs, by.x = "HomeTeam", by.y = "team_name", all.x = T)
names(ACCGames)[which(names(ACCGames) == "team_id")] = "HomeTeamID"
gameIDRefs <- unique(ACCGames[,c("HomeTeam", "AwayTeam", "GameDate", "NeutralSite", "AwayTeamID", "HomeTeamID")])
gameIDRefs <- arrange(gameIDRefs, GameDate, HomeTeamID, AwayTeamID)
gameIDRefs$gameID <- c(1:nrow(gameIDRefs))
ACCGames <- merge(ACCGames, gameIDRefs, by = c("HomeTeam", "AwayTeam", "GameDate", "NeutralSite", "AwayTeamID", "HomeTeamID"), all.x = T)
```


After creating both team ID and game ID reference tables, I did some basic feature engineering to calculate total rebounding percentage, offensive rebounding percentage, and defensive rebounding percentage for each team in each game ($TRBPct$, $ORBPct$, and $DRBPct$ respectively). I then went on to calculate field goal percentage, three point field goal percentage, free throw attempt rate, three point field goal attempt rate, and defensive activity ((steals + rebounds)/personal fouls) both for each team in each game and for each team on a per-game level. 


```{r}
####Feature Engineering####
##Calculate Rebound Pct
ACCGames$HomeRebOpps <- (ACCGames$AwayFTA + ACCGames$HomeFTA) * 0.44 + (ACCGames$AwayFGA + ACCGames$HomeFGA) + (ACCGames$AwayBLK + ACCGames$HomeBLK)
ACCGames$AwayRebOpps <- (ACCGames$AwayFTA + ACCGames$HomeFTA) * 0.44 + (ACCGames$AwayFGA + ACCGames$HomeFGA) + (ACCGames$AwayBLK + ACCGames$HomeBLK)
ACCGames$HomeOffRebOpps <- (ACCGames$HomeFTA) * 0.44 + ACCGames$HomeFGA + ACCGames$AwayBLK
ACCGames$AwayOffRebOpps <- (ACCGames$AwayFTA) * 0.44 + ACCGames$AwayFGA + ACCGames$HomeBLK
ACCGames$AwayDefRebOpps <- (ACCGames$HomeFTA) * 0.44 + ACCGames$HomeFGA + ACCGames$AwayBLK
ACCGames$HomeDefRebOpps <- (ACCGames$AwayFTA) * 0.44 + ACCGames$AwayFGA + ACCGames$HomeBLK
ACCGames$HomeTRBPct <- ACCGames$HomeRebounds/ACCGames$HomeRebOpps
ACCGames$AwayTRBPct <- ACCGames$AwayRebounds/ACCGames$AwayRebOpps
ACCGames$HomeORBPct <- ACCGames$HomeORB/ACCGames$HomeOffRebOpps
ACCGames$AwayORBPct <- ACCGames$AwayORB/ACCGames$AwayOffRebOpps
ACCGames$HomeDRBPct <- ACCGames$HomeDRB/ACCGames$HomeDefRebOpps
ACCGames$AwayDRBPct <- ACCGames$AwayDRB/ACCGames$AwayDefRebOpps
  
##Create Per Game Averages and SDs for Each Team  
awayVars <- names(ACCGames)[which(grepl("Away", names(ACCGames)) == T)]
homeVars <- names(ACCGames)[which(grepl("Home", names(ACCGames)) == T)]
homeTeamDT <- ACCGames[,c("GameDate", "NeutralSite", "gameID", homeVars), with = F]
awayTeamDT <- ACCGames[,c("GameDate", "NeutralSite", "gameID", awayVars), with = F]
names(homeTeamDT) <- gsub("Home", "", names(homeTeamDT))
names(awayTeamDT) <- gsub("Away", "", names(awayTeamDT))
homeTeamDT$HomeAway <- rep(1, nrow(homeTeamDT))
awayTeamDT$HomeAway <- rep(0, nrow(awayTeamDT))
fullTeamDT <- rbind(homeTeamDT, awayTeamDT)
fullTeamDT <- setDT(arrange(fullTeamDT, gameID, -HomeAway))
names(fullTeamDT) <- gsub("3FGM", "FG3M", names(fullTeamDT))
names(fullTeamDT) <- gsub("3FGA", "FG3A", names(fullTeamDT))
fullTeamDT$FGPct <- fullTeamDT$FGM/fullTeamDT$FGA
fullTeamDT$FG3Pct <- fullTeamDT$FG3M/fullTeamDT$FG3A
fullTeamDT$FTPct <- fullTeamDT$FTM/fullTeamDT$FTA
fullTeamDT$FTARate <- fullTeamDT$FTA/fullTeamDT$FGA
fullTeamDT$FG3ARate <- fullTeamDT$FG3A/fullTeamDT$FGA
fullTeamDT$DefActivity <- (fullTeamDT$STL + fullTeamDT$BLK)/fullTeamDT$Fouls
teamVars <- names(fullTeamDT)[which(names(fullTeamDT) %nin% c("GameDate", "NeutralSite", "gameID", "Team", "TeamID", "HomeAway"))]
fullTeamDT <- fullTeamDT[, paste0(teamVars, "Diff") := lapply(.SD, function(x) c(-1,1)*diff(x)), by = "gameID", .SDcols = c(teamVars)]
teamVars <- c(teamVars, paste0(teamVars, "Diff"))
fullTeamDT <- fullTeamDT[, paste0(teamVars, "PerGame") := lapply(.SD, mean, na.rm = T), by = "TeamID", .SDcols = c(teamVars)]
fullTeamDT <- fullTeamDT[, paste0(teamVars, "SD") := lapply(.SD, sd, na.rm = T), by = "TeamID", .SDcols = c(teamVars)]
teamVars <- c(teamVars, paste0(teamVars, "PerGame"), paste0(teamVars, "SD"))
ACCGamesFull <- unique(ACCGames[,c("HomeTeam", "AwayTeam", "GameDate", "NeutralSite", "AwayTeamID", "HomeTeamID", "gameID")])
ACCGamesFull <- merge(ACCGamesFull, fullTeamDT[,c("TeamID", "gameID", teamVars), with = F], by.x = c("gameID", "HomeTeamID"), by.y = c("gameID", "TeamID"), all.x = T)
names(ACCGamesFull)[which(names(ACCGamesFull) %in% teamVars)] <- paste0("Home", teamVars)
ACCGamesFull <- merge(ACCGamesFull, fullTeamDT[,c("TeamID", "gameID", teamVars), with = F], by.x = c("gameID", "AwayTeamID"), by.y = c("gameID", "TeamID"), all.x = T)
names(ACCGamesFull)[which(names(ACCGamesFull) %in% teamVars)] <- paste0("Away", teamVars)
```


After concluding feature engineering, I populated a binary team matrix with 1's for the home team of each game and -1's for the away team of each game. $Margin$ is the target variable and is defined as the score margin for the home team.  

```{r}
####Create Model Matrix with Teams, Home/Away, and Margin####
##Home and Away
HomeMat <- model.matrix(~ as.character(HomeTeamID) + 0, ACCGamesFull)
AwayMat <- model.matrix(~ as.character(AwayTeamID) + 0, ACCGamesFull)
colnames(HomeMat) <- paste0("Home_ID", gsub(" |as.character[(]HomeTeamID[)]|[(]|[)]", "", colnames(HomeMat)))
colnames(AwayMat) <- paste0("Away_ID", gsub(" |as.character[(]AwayTeamID[)]|[(]|[)]", "", colnames(AwayMat)))
TeamMat <- cbind(HomeMat, -AwayMat)

##Margin
Margin <- ACCGamesFull$HomeScore - ACCGamesFull$AwayScore
```



Finally, I created the total prediction matrix $x$ by binding the binary team matrix with a constant, the $NeutralSite$ indicator variable, and all of the variables listed below in the $impVars$ vector (both per-game features and individual game features were included for these). 

I used the cv.glmnet function to standardize the data and run cross validation before determining an optimal lambda variable. As seen below, each team's predictive score is stotred in $marginResults$.
```{r}
####CV Ridge Regression####
impVars <- c("FGPct", "FG3Pct", "FTPct", "FG3ARate", "FTARate", "DefActivity", "TRBPct", "ORBPct", "DRBPct")
impVars <- c(impVars, paste0(impVars, "PerGame"))
x <- cbind(Constant = 1, TeamMat, NeutralSite = ACCGamesFull$NeutralSite, as.matrix(ACCGamesFull[,paste0(c("Home", "Away"), impVars), with = F]))
y <- Margin
lambdas <- seq(50, 8000, 50)
fit <- glmnet(x, y, alpha = 1, lambda = lambdas)
cv_fit <- cv.glmnet(x, y, alpha = 0, lambda = lambdas, standardize = T, standardize.response = T)
opt_lambda <- cv_fit$lambda.min
fit <- cv_fit$glmnet.fit
allBetas <- cv_fit$glmnet.fit$beta[,which(cv_fit$glmnet.fit$lambda == opt_lambda)]
allBetas <- allBetas[which(grepl("_ID", names(allBetas)) == T)]
marginResults <- as.data.frame(do.call(rbind, strsplit(names(allBetas), "_ID")))
names(marginResults) <- c("margin_type", "team_id")
marginResults$beta <- allBetas
marginResults <- merge(marginResults, teamIDRefs, by = "team_id", all.x = T)
marginResults <- spread(marginResults, margin_type, beta)
marginResults$Total <- marginResults$Home + marginResults$Away
marginResults <- arrange(marginResults, -Total)
print(marginResults)
```


